---
title: "07_changes"
format:
  html:
    code-fold: true
---


Here we look a changes over time for individual species as well as for the aggregate.  We bin abundance into a hexagonal mesh covering the area, compute long term means and the examine trends over time relative to each long-term mean. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("setup.R")
```


```{r load_data}
x = read_ecomon_spp(form = 'sf') |> ecomon_to_long()
h = read_hexgrid()
coast = read_coast()
plot(sf::st_geometry(x), pch = ".", axes = TRUE, reset = FALSE, main = "All records")
plot(sf::st_geometry(h), border = "gray", add = TRUE)
plot(coast, col = "orange", add = TRUE)
```


# Summaries across all time

Here we create long term total and mean per hexagon

```{r total_abundance}
totabund = tally_abundance(x, h, fun = sum, na.rm = TRUE, shape = "long") |>
  write_abundance(name = "all_total_abund_hex")
```

```{r mean_abundance}
meanabund = tally_abundance(x, h, fun = mean, na.rm = TRUE, shape = "long") |>
 write_abundance(name = "all_mean_abund_hex")
```


We'll read back in the mean abundance in wide format, and plot that.

```{r plot_each}
meanabund = meanabund |> wider_abundance()
nbreaks = 10
plot(meanabund,
     nbreaks = nbreaks,
     pal = RColorBrewer::brewer.pal(nbreaks-1, "Blues"))
```
  
  
    