---
title: "07_changes"
format:
  html:
    code-fold: true
---


Here we look a changes over time for individual species as well as for the aggregate.  We bin abundance into a hexagonal mesh covering the area, compute long term means and the examine trends over time relative to each long-term mean. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("setup.R")
```


```{r load_data}
decades = seq(as.Date("1970-01-01"), as.Date("2020-01-01"), by = "10 years")
x = read_ecomon_spp(form = 'sf') |> ecomon_to_long()
sf_col = attr(x, "sf_column")
x = x |>
  dplyr::mutate(year = format(date, "%Y"),
                month = format(date, "%b") |> factor(levels = month.abb),
                decade = decades[findInterval(.data$date, decades)],
                .before = dplyr::all_of(sf_col))
h = read_hexgrid()
coast = read_coast()
plot(sf::st_geometry(x), pch = ".", axes = TRUE, reset = FALSE, main = "All records")
plot(sf::st_geometry(h), border = "gray", add = TRUE)
plot(coast, col = "orange", add = TRUE)
```


# Summaries across all time

Here we create long term total and mean per hexagon

```{r total_abundance}
totabund = tally_abundance(x, h, fun = sum, na.rm = TRUE, shape = "long") |>
  wider_abundance()
nbreaks = 10
plot(totabund,
     nbreaks = nbreaks,
     pal = RColorBrewer::brewer.pal(nbreaks-1, "Greens"))
```


# Mean abudance

```{r plot_each}
mabund = tally_abundance(x, h, fun = mean, na.rm = TRUE, 
                         shape = "long") |>
  wider_abundance()
nbreaks = 10
plot(mabund,
     nbreaks = nbreaks,
     pal = RColorBrewer::brewer.pal(nbreaks-1, "Blues"))
```
  
# Decadal abundance

Here we look at decadal abundance departures for the long term mean. 

```{r}
y = x |>
  group_by(name, decade) |>
  group_map(
    function(tbl, key){
      tally_by_poly(tbl, h, fun = mean, na.rm = TRUE, shape = "long") |>
        dplyr::mutate(name = key$name, decade = key$decade, .before = 1)
    }
  ) |>
  dplyr::bind_rows()
```

```{r}
ta = tally_abundance(x, h, fun = mean, na.rm = TRUE, shape = "long")
```


```{r}
dy = y |>
  dplyr::group_by(name, decade) |>
  dplyr::group_map(
    function(tbl, key){
      d = tbl$value - dplyr::filter(ta, name == key$name) |> dplyr::pull(value)
      dplyr::mutate(tbl, 
                    name = key$name,
                    dvalue = d, 
                    decade = key$decade,
                    .after = dplyr::all_of("value"))  
    } ) |>
  dplyr::bind_rows()
```


```{r}

ggplot(dy) +
  geom_sf(color = "white", aes(fill = dvalue)) +
  geom_sf(data = coast) + 
  scale_fill_distiller(palette = "Spectral", na.value = NA, type = "div") +
  facet_grid(name ~ decade)
```

# Monthly climatologies

Here we compute means by month for each species (and the total.)  

```{r}
ma = x |>
  group_by(name, month) |>
  group_map(
    function(tbl, key){
      tally_by_poly(tbl, h, fun = mean, na.rm = TRUE, shape = "long") |>
        dplyr::mutate(name = key$name, month = key$month, .before = 1)
    }
  ) |>
  dplyr::bind_rows()
```

```{r}
y = x |>
  group_by(name, decade, month) |>
  group_map(
    function(tbl, key){
      tally_by_poly(tbl, h, fun = mean, na.rm = TRUE, shape = "long") |>
        dplyr::mutate(name = key$name, decade = key$decade, month = key$month, .before = 1)
    }
  ) |>
  dplyr::bind_rows()
```

And here we find the departure from the monthly mean for each decade.
```{r}
dy = y |>
  dplyr::group_by(name, decade, month) |>
  dplyr::group_map(
    function(tbl, key){
      d = tbl$value - dplyr::filter(ma, name == key$name, month == key$month) |> 
        dplyr::pull(value)
      dplyr::mutate(tbl, 
                    name = key$name,
                    dvalue = d, 
                    month = key$month,
                    decade = key$decade, 
                    .after = dplyr::all_of("value"))  
    } ) |>
  dplyr::bind_rows()
```

We plot these in a series by species.
```{r}
spp = count(dy, name) |> pull(var = 1)
sppname = sub("_10m2", "", spp, fixed = TRUE)
gg = lapply(seq_along(spp),
  function(i){
    ggplot(data = filter(dy, name == spp[i])) +
      geom_sf(color = "white", aes(fill = dvalue)) +
      geom_sf(data = coast) + 
      scale_fill_distiller(palette = "Spectral", na.value = NA, type = "div") +
      labs(fill = sppname[i]) + 
      facet_grid(decade ~ month) + 
      theme(axis.text = element_blank())
  })
for (g in gg) g
```


